The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching

### 2016-09-06 article.tex
Updated the performance evaluation of molecular formula generators

  The CDK contains two algorithms for the decomposition of mass ranges into
possible elemental formulas. For most inputs, a Round Robin algorithm,
originally developed for the SIRIUS metabolite identification
tool~\cite{Bocker2009}, is used. The algorithm discretizes the real-value mass
decomposition problem into an integer-value knapsack
problem~\cite{Martello1990}. It first computes a dynamic programming table and
then backtracks it to generate matching formulas~\cite{Duehrkop2013,
Boecker2008}. Data for the Round Robin algorithm is stored in an extended
residue table~\cite{Bocker2005}, resulting in a low memory footprint of several
kilobytes. For certain problem instances, such as very large mass values (above
$400\,000$ Da) or mass range span larger than $1$ Da, the Round Robin algorithm
is not applicable and CDK falls back to an optimized full enumeration search
method, originally developed as part of the MZmine~2 framework for mass
spectrometry data processing~\cite{Pluskal2012, Pluskal2010}.

@@

   To evaluate the performance of the CDK molecular formula generator, we
compared its runtimes on a range of inputs and compared them to the runtimes of
the classic, full enumeration-based HR2 formula generator~\cite{Kind2007} and
those of a recently developed Parallel Formula Generator (PFG)~\cite{Zhang2016}
(Table~1\ref{tab:formula_generators}). Although for small mass values the PFG
algorithm achieved shorter runtimes, for larger masses the Round Robin
algorithm implemented in CDK clearly outperformed the two other methods.

@@

  \subsection*{Table 1 - Evaluation of molecular formula generators.}
  \label{tab:formula_generators}
  The resulting formula counts and runtimes of the HR2, PFG, and CDK chemical
formula generators based on input mass and mass tolerance values. Formulas were
generated using chemical elements C,H,N,O,P,S without bounds (the allowed atom
count was set to 0--10000 for each element). All heuristic filtering rules were
disabled for the purpose of the evaluation. The slight differences in the
number of generated formulas were caused by different isotope masses embedded
in each software and/or by rounding errors during calculation. The runtimes are
average values from three independent runs performed on three different 16-core
Intel Xeon 2.2~GHz CPU workstations equipped with 128~GB RAM, running Ubuntu
Linux version 12.04.5 LTS and OpenJDK Java runtime version 1.7.0\_101.
  \vskip 1\baselineskip

    \begin{minipage}{1\textwidth}
    \centering
    \begin{tabular}{llllllll}
	\multirow{2}{*}{ Mass (Da) } & \multirow{2}{*}{ Tolerance ($\pm$ Da) } & \multicolumn{3}{c}{ \# of generated formulas } & \multicolumn{3}{c}{ Runtime (s) } \\
	& & HR2 & PFG & CDK & HR2 & PFG & CDK \\
	$100$ & $0.001$ & $1$ & $1$ & $1$ & $0.10$ & $\mathbf{0.06}$ & $0.49$ \\
	$100$ & $0.01$ & $18$ & $18$ & $18$ & $0.03$ & $\mathbf{0.01}$ & $0.52$ \\
	$500$ & $0.001$ & $430$ & $430$ & $430$ & $0.51$ & $\mathbf{0.09}$ & $0.67$ \\
	$500$ & $0.01$ & $4\,265$ & $4\,265$ & $4\,264$ & $0.55$ & $\mathbf{0.39}$ & $0.96$ \\
	$1\,000$ & $0.001$ & $8\,572$ & $8\,572$ & $8\,568$ & $16.5$ & $1.5$ & $\mathbf{1.2}$ \\
	$1\,000$ & $0.01$ & $85\,542$ & $85\,542$ & $85\,537$ & $16.8$ & $5.4$ & $\mathbf{3.7}$ \\
	$2\,000$ & $0.001$ & $226\,346$ & $226\,346$ & $226\,323$ & $1156.5$ & $58.9$ & $\mathbf{8.6}$ \\
	$2\,000$ & $0.01$ & $2\,263\,410$ & $2\,263\,410$ & $2\,263\,409$ & $932.5$ & $156.3$ & $\mathbf{69.7}$ \\
	$3\,000$ & $0.001$ & $1\,647\,132$ & $1\,647\,132$ & $1\,647\,118$ & $9\,667.9$ & $603.5$ & $\mathbf{59.3}$ \\
	$3\,000$ & $0.01$ & $16\,471\,338$ & $16\,471\,338$ & $16\,471\,373$ & $9\,694.5$ & $1\,303.6$ & $\mathbf{505.7}$ \\
    \end{tabular}
    \end{minipage}

>>>>

  The CDK contains two algorithms for the decomposition of mass ranges into
possible elemental formulas. For most inputs, a Round Robin algorithm,
originally developed for the SIRIUS metabolite identification
tool~\cite{Bocker2009}, is used. The algorithm discretizes the real-value mass
decomposition problem into an integer-value knapsack
problem~\cite{Martello1990}. It first computes a dynamic programming table and
then backtracks it to generate matching formulas~\cite{Duehrkop2013,
Boecker2008}. Data for the Round Robin algorithm is stored in an extended
residue table~\cite{Bocker2005}, resulting in a low memory footprint of several
kilobytes. For certain problem instances, such as very large mass values (above
$400\,000$ Da) or mass range span larger than $1$ Da, the Round Robin algorithm
is not suitable and CDK falls back to an optimized full enumeration search
method, originally developed as part of the MZmine~2 framework for mass
spectrometry data processing~\cite{Pluskal2012, Pluskal2010}.

@@

   To evaluate the performance of the CDK molecular formula generator, we
compared its runtimes to those of the classic, full enumeration-based HR2
formula generator~\cite{Kind2007} and those of a recently developed Parallel
Formula Generator (PFG)~\cite{Zhang2016} (Table~1\ref{tab:formula_generators}).
As inputs, we used two sets of $10\,000$ small ($< 500$~Da) and $20$ large ($>
1\,500$ and $< 3\,500$~Da) molecular mass values downloaded from the Global
Natural Products Social Molecular Networking database~\cite{wang2016}. The mass
tolerance was set to 0.001 or 0.01 Da. The CDK's Round-Robin formula generator
outperformed the other methods in all cases, despite running in a single thread
(PFG utilizes multiple threads).

@@

  \subsection*{Table 1 - Evaluation of molecular formula generators.}
  \label{tab:formula_generators}
  The resulting formula counts and runtimes of the HR2, PFG, and CDK chemical
formula generators on two different inputs with two different mass tolerance
settings. For the set of small masses, $10\,000$ mass values in the range of
$0$--$500$ Da were randomly selected from the Global Natural Products Social
Molecular Networking database~\cite{wang2016}. For the set of large masses, $20$
mass values in the range of $1\,500$--$3\,500$ Da were randomly selected from the same
database. Formulas were generated using chemical elements C,H,N,O,P,S without
bounds (the allowed atom count was set to $0$--$10\,000$ for each element). All
heuristic filtering rules were disabled for the purpose of the evaluation. The
slight differences in the number of generated formulas were caused by different
isotope masses embedded in each software and/or by rounding errors during
calculation. The runtimes are average values from three independent runs
performed on three different 16-core Intel Xeon 2.9~GHz CPU workstations
equipped with 189~GB RAM, running Ubuntu Linux version 12.04.5 LTS and OpenJDK
Java runtime version 1.7.0\_101.
  \vskip 1\baselineskip

    \begin{minipage}{1\textwidth}
    \centering
    \begin{tabular}{llllllll}
	\multirow{2}{*}{ Input } & \multirow{2}{*}{ Mass tolerance ($\pm$ Da) } & \multicolumn{3}{c}{ \# of generated formulas } & \multicolumn{3}{c}{ Runtime (s) } \\
	& & HR2 & PFG & CDK & HR2 & PFG & CDK \\
	$10\,000$ small masses & $0.001$ & $616\,846$ & $616\,846$ & $616\,843$ & $669$ & $168$ & $\mathbf{41}$ \\
	$10\,000$ small masses & $0.01$ & $6\,163\,303$ & $6\,163\,302$ & $6\,163\,326$ & $689$ & $501$ & $\mathbf{212}$ \\
	$20$ large masses & $0.001$ & $4\,912\,939$ & $4\,912\,939$ & $4\,912\,904$ & $26\,370$ & $1\,292$ & $\mathbf{177}$ \\
	$20$ large masses & $0.01$ & $49\,128\,811$ & $49\,128\,810$ & $49\,128\,815$ & $26\,587$ & $3\,406$ & $\mathbf{1\,580}$ \\
    \end{tabular}
    \end{minipage}
